---
title: "Your Paper Title Here"
author: "Your Name(s)"
date: Date
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
```

# Abstract
Write your abstract here.

# Keywords
*Likelihood Analysis*, *Bayesian Statistics*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background
Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

# Statistical Methods

## Model
Describe the statistical model used.

## Likelihood Inference
Detail the likelihood approach.

## Bayesian Inference
$$T\sim Binom(170, \pi)$$

```{r}
t <- 8
n <- 170
mle <- 154/162
mle <- (n-2*t)/(n-t)
se <- 1/sqrt(8/((1-mle)^2)-170/((2-mle)^2))
z <- qnorm(0.975)

#wrong i think
lower <- mle - z * se
upper <- mle + z * se

cat("[",lower,", ",upper,"]",sep="")

#second deriv is positive??
```
$$f(t)={170 \choose t}\pi^t(1-\pi)^{170-t} \ t=0,1,2,...,170$$
$$\pi=\frac{1-\psi}{2-\psi}$$
$$f(t)={170 \choose t}(\frac{1-\psi}{2-\psi})^{t}(\frac{1}{2-\psi})^{170-t}$$
$$L(\psi)={170 \choose 8}(\frac{1-\psi}{2-\psi})^8(\frac{1}{2-\psi})^{162}$$
$$\ell(\psi)=ln{170 \choose 8}+8ln(1-\psi)-8ln(2-\psi)+162ln1-162ln(2-\psi)$$
$$\frac{d}{d\psi}\ell(\psi)=0-\frac{8}{1-\psi}+\frac{8}{2-\psi}+0+\frac{162}{2-\psi}=0$$
$$\frac{170}{2-\psi}=\frac{8}{1-\psi}$$
$$170-170\psi=16-8\psi$$
$$154=162\psi$$
$$\widehat{\psi}_0^{mle}=\frac{154}{162}\approx 0.9506$$
## Second Derivative Test

$$\frac{d^2}{d\psi^2}\ell(\psi)=-\frac{8}{(1-\psi)^2}+\frac{170}{(2-\psi)^2}$$

```{r}
log_lik <- function(psi){
  log(choose(170,8))+8*log(1-psi)-8*log(2-psi)-162*log(2-psi)
}

second_deriv <- function(psi){
  -8/((1-psi)^2) + 170/((2-psi)^2)
}

ggplot() +
  geom_function(fun=second_deriv, xlim=c(-100,1)) +
  ylim(c(-1,1))
#positive if psi is positive
```

## Bootstrapping

```{r}
set.seed(414)

B <- 10000
sample_size <- 15
pi <- (1-mle)/(2-mle)

boot_df <- as.data.frame(replicate(n=B, expr={x <- rbinom(n=1,size=170, prob=pi)
                     (170 - 2*x)/(170-x)
                    }))
names(boot_df) <- c("est")
boot_df

lwr <- quantile(boot_df$est, 0.025)
upr <- quantile(boot_df$est, 0.975)

cat("[",lwr,", ",upr,"]",sep="")
```

##Likelihood ratio significance test
```{r}
# H0: psi is 0.3 . H1: otherwise
null <- 0.3
W <- 2*log_lik(mle)-2*log_lik(null)
pchisq(W, df=1, lower.tail=F)
```

##Empirical p value
```{r}
set.seed(414)

unifs <- replicate(n=B, expr={x <- sample(size=1, 1:170, replace=T)
                    mle <- (170-2*x)/(170-x)
                    W <- 2*log_lik(mle)-2*log_lik(null)
                    })

sum(unifs>W, na.rm=T)
```

# Results
Present your findings. 

# Discussion / Conclusion
Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from https://www.example.com.

# Appendix

## Code
```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs
If applicable, include detailed mathematical derivations or additional theoretical explanations.
