---
title: "Your Paper Title Here"
author: "Alison , Katie, Viki"
date: Date
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
```

# Abstract
Write your abstract here.

# Keywords
*Keyword 1*, *Keyword 2*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background
In December 2020, Pfizer and BioNTech began distributing their two dose vaccine for COVID-19. 

The data comes from blah blah blah

make a segmented barplot and table of data (side by side)

From the initial data we have these statistics. 

```{r intro, eval=TRUE, echo=FALSE}
library(knitr)
library(ggplot2)

#data table
vaccine_table <- data.frame( Group = c("BNT162b2", "Placebo", "Total"), Cases = c(8, 162, 170), Sample_size = c(17411, 17511, 34922))

kable(vaccine_table, caption = "Table 1: Vaccine Efficacy against Covid-19 at least 7 days after second dose in patients without evidence of infection")

# summary stats table
summary_table <- data.frame( Statistic = c( "$\\hat{\\pi}_v$ (vaccine infection probability)", "$\\hat{\\pi}_p$ (placebo infection probability)", "$RR$ (relative risk)", "$\\psi$ (vaccine efficacy)"),
  Formula = c("$8/17411$", "$162/17511$", "$\\hat{\\pi}_v/\\hat{\\pi}_p$", "$1 - \\hat{\\pi}_v/\\hat{\\pi}_p$"),
  Estimate = c(8/17411, 162/17511, (8/17411)/(162/17511), 1 - (8/17411)/(162/17511)))

kable(summary_table, caption = "Estimated Risk and Vaccine Efficacy", digits = 6)

cases <- data.frame( Group = c("BNT162b2", "Placebo"), Cases = c(8, 162) ) 
ggplot(cases, aes(x = "Cases", y = Cases, fill = Group)) + geom_bar(stat = "identity") + labs(x = "", y = "Number of Cases", title= "COVID-19 Cases by Treatment Group")
```



# Statistical Methods

## Model
  Instead of looking at two independent binomial models, since the trial ran until a predetermined 170 observations of COVID-19 infections across both groups was observed, we can use a single binomial random variable as our statistical model (refer to appendix for proof). We can denote random variable T as the number in the vaccine group from the total 170 observations of infection. T can be modeled as (refer to appendix for proof): 
$$T \approx Binom(n=170, \pi)$$
  Here $t_{obs}=8$ and $\pi=P(\text{case occured in vaccine group})$. We can write $\pi$ as $\pi=\frac{\pi_vn_v}{\pi_vn_v+\pi_pn_p}=\frac{\pi_v}{\pi_v+\pi_p}$ since the sample sizes of the group receiving the placebo and the group receiving the vaccine are roughly the same ($n_p\approx n_v$). 

  Our parameter of interest is vaccine efficacy. We can rewrite this, $\psi$, in terms of $\pi$ to get (refer to appendix for proof): $$\psi = \frac{1-2\pi}{1-\pi} \qquad -\infty < \psi < 1$$

  With this we can also find $\pi$ in terms of $\psi$ for later use in our Frequentist and Bayesian analysis of our data. We get (refer to appendix for proof): $$\pi=\frac{1-\psi}{2-\psi}\qquad 0 < \pi < 1$$

  Because the FDA requires at least $30\%$ efficacy for a new therapy to be approved, the hypothesis we are interested in testing for both approaches is $$H_0: \psi=0.3 \qquad   H_1:\psi>0.3$$


## Likelihood Inference
Detail the likelihood approach.

## Bayesian Inference
Detail the Bayesian approach.

# Results
Present your findings. 

# Discussion / Conclusion
Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from https://www.example.com.

# Appendix

## Code
```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs

#### Proof that $T \approx Binom$: 

We start with our two independent binomial models $X \backsim Binom(n_v,\pi_v)$ and $Y\backsim Binom(n_p, \pi_p)$. Here, X is the number of COVID-19 cases in the vaccine group and Y is the number of COVID-19 cases in the placebo group. Restricting our attention to only the COVID-19 cases, we can define the random variable $T$ as the number of COVID-19 cases in the vaccine group given that the total number of COVID-19 cases in both groups is 170. In other words, $T$ is the value of $X$ given $X+Y=170$.

So, we can find the PMF of $T$ by:
$$P(T=t) = P(X=t|X+Y = n) \qquad t=0,1,\dots, n$$
$$=\frac{P(X=t \cap X+Y=n)}{P(X+Y = n)}$$
$$=\frac{P(X=t \cap Y=n-t)}{P(X+Y = n)}$$
Since X and Y are independent, $$=\frac{P(X=t) \times P(Y=n-t)}{P(X+Y = n)}$$

We'll come back to this, but for now, referring back to table 2 in the introduction, we can see that $n_v$ and $n_p$ are very large and $\pi_v$ and $\pi_p$ are very small. So, these two binomial models can be approximated by 2 independent poisson distribution: $X\approx Pois(\lambda_v)$ and $Y\approx Pois(\lambda_p)$ where $\lambda_v = n_v\pi_v$ and $\lambda_p=n_p\pi_p$. 

For a poisson distribution $P(X=x)=\frac{\lambda^x e^{-\lambda}}{x!}$ So, 
$$P(X=t)=\frac{\lambda_v^t e^{-\lambda_v}}{t!} \qquad P(Y=n-t)=\frac{\lambda_p^{n-t} e^{-\lambda_p}}{(n-t)!}$$
Since the two poisson distributions are still independent of each other, we can add them to get $X+Y \backsim Pois(\lambda_v + \lambda_p)$. So,
$$P(X+Y=n)=\frac{(\lambda_p+\lambda_v)^{n} e^{-\lambda_p-\lambda_v}}{n!}$$
Plugging these into the PMF we found for $T$, we get: 
$$P(T=t)=\frac{\frac{\lambda_v^t e^{-\lambda_v}}{t!} \times \frac{\lambda_p^{n-t} e^{-\lambda_p}}{(n-t)!}}{\frac{(\lambda_p+\lambda_v)^{n} e^{-\lambda_p-\lambda_v}}{n!}}$$
$$P(T=t)=\frac{n!}{t!(n-t)!} \times \frac{\lambda_v^t\lambda_p^{n-t} e^{-\lambda_p}e^{-\lambda_v}}{(\lambda_p+\lambda_v)^{n} e^{-\lambda_p-\lambda_v}}$$
$$P(T=t)=\frac{n!}{t!(n-t)!} \times \frac{\lambda_v^t\lambda_p^{n-t}}{(\lambda_p+\lambda_v)^{n}}$$
$$P(T=t)=\frac{n!}{t!(n-t)!} \times \frac{\lambda_v^t\lambda_p^{n-t}}{(\lambda_p+\lambda_v)^{t}(\lambda_p+\lambda_v)^{n-t}}$$

$$P(T=t)=\frac{n!}{t!(n-t)!} \times \frac{\lambda_v^t}{(\lambda_p+\lambda_v)^{t}} \times \frac{\lambda_p^{n-t}}{(\lambda_p+\lambda_v)^{n-t}}$$
$$\text{The choose function: }\frac{n!}{t!(n-t)!} = {n \choose t}$$
$$P(T=t)={n \choose t} \times (\frac{\lambda_v}{(\lambda_p+\lambda_v)})^{t} \times (\frac{\lambda_p}{(\lambda_p+\lambda_v)})^{n-t}$$
We can rewrite the third term $(\frac{\lambda_p}{(\lambda_p+\lambda_v)})^{n-t}$ as $$(1-\frac{\lambda_v}{(\lambda_p+\lambda_v)})^{n-t}$$

$$P(T=t)={n \choose t} \times (\frac{\lambda_v}{(\lambda_p+\lambda_v)})^{t} \times (1-\frac{\lambda_v}{(\lambda_p+\lambda_v)})^{n-t}$$
And yay! Looking at this closely, this is the binomial PMF for $Binom(n,\frac{\lambda_v}{(\lambda_p+\lambda_v)})$. So, we can say $$T\approx Binom(n,\frac{\lambda_v}{(\lambda_p+\lambda_v)})$$ since $\lambda_v = n_v\pi_v$ and $\lambda_p=n_p\pi_p$, we can say $$T \approx Binom(n,\frac{n_v\pi_v}{n_p\pi_p+n_v\pi_v}) \qquad \text{where } n=170$$

Finally, knowing that $n=170$ and since $n_v$ and $n_p$ are really large, we can say for $\pi=\frac{\pi_v}{\pi_p+\pi_v}$ $$T\approx Binom(170, \pi)$$






#### Proof of $\psi = \frac{1-2\pi}{1-\pi}$:

The definition of vaccine efficacy ($\psi$) is: $\psi = 1-\frac{\pi_v}{\pi_p}$, we can rewrite this in terms of $\pi$ using:
$$\pi=\frac{\pi_v}{\pi_v+\pi_p}$$
$$\pi(\pi_v+\pi_p)=\pi_v$$
$$\pi\pi_v+\pi\pi_p=\pi_v$$
$$\pi\pi_p=\pi_v-\pi\pi_v$$
$$\pi\pi_p=\pi_v(1-\pi)$$
$$\frac{\pi_v}{\pi_p}=\frac{\pi}{1-\pi}$$

Plugging $\frac{\pi_v}{\pi_p}=\frac{\pi}{1-\pi}$ into $\psi = 1-\frac{\pi_v}{\pi_p}$, we get: $$\psi = 1-\frac{\pi}{1-\pi}$$ $$\psi = \frac{1-\pi}{1-\pi}-\frac{\pi}{1-\pi}$$ $$\boxed{\psi = \frac{1-2\pi}{1-\pi} \qquad -\infty < \psi < 1}$$


#### Proof of $\pi=\frac{1-\psi}{2-\psi}$:

Using the equation we deduced in the previous proof we can rearrange $\psi = \frac{1-2\pi}{1-\pi}$ to create an equation for $\pi$ in terms of $\psi$. 
$$\psi = \frac{1-2\pi}{1-\pi}$$
$$\psi(1-\pi) = 1-2\pi$$
$$\psi-\psi\pi = 1-2\pi$$
$$2\pi-\psi\pi = 1-\psi$$
$$\pi(2-\psi) = 1-\psi$$
We finally get $$\boxed{\pi = \frac{1-\psi}{2-\psi} \qquad 0 < \pi < 1}$$

